Refinement of Thermostated Molecular Dynamics via Backward Error Analysis

### 2018-07-20 rigid_bodies.tex
small changes

It seems then reasonable to argue that any thermostat should regulate the shadow kinetic energy instead of the specified one.
However, the theory underlying the concept of shadow Hamiltonian is only applicable to symplectic integrators.
In this paper, we devise a novel procedure that partially overcomes this limitation.
It consists in using a splitting-based numerical integrator that entails a Hamiltonian component and carrying out backward error analysis out of this NVE part only.
This allows us to obtain, with minimal computational overhead, a \textit{refined kinetic energy} and a \textit{refined Hamiltonian} as suitable estimators of their shadow counterparts.
By recasting the thermostat equations with this refined kinetic energy, we are able to predict its equipartition and to reproduce a canonical ensemble consistent with the refined Hamiltonian.
Finally, a reweighting procedure provides us with ensemble averages which are, instead, consistent with the specified one.
Although we will show the detailed derivation for the particular case of the Nos\'{e}-Hoover chain thermostat, it should be emphasized that our new approach is not limited to deterministic thermostats. 
It is straightforward to use it, for instance, in conjunction with global stochastic thermostats such as those of Bussi \textit{et al}. \cite{Bussi_2007} and Leimkuhler \textit{et al}. \cite{Leimkuhler_2009}, for instance.

Our new method is particularly appealing for rigid-body MD, primarily because its numerical stability extends up to time steps large enough to yield pronounced discretization errors. Rigid-body MD is relevant not only because some molecules are usually modeled as rigid, such as in the ubiquitous case of water \cite{Jorgensen_1983}, but also because it can provide a coarse-graining strategy for molecules that can be designed as collections of interconnected bodies, such as proteins and nanoparticles \cite{Miller_2002, Knorowski_2012, Patra_2013}.
In fact, the approach of Kamberaj \textit{et al}. \cite{Kamberaj_2005}, which is implemented in software packages such as HOOMD-\textit{blue}\cite{Anderson_2008} and LAMMPS\cite{Plimpton_1995}, has been applied to simulate small molecules \cite{Geiger_2013, Aimoli_2014, Aimoli_2014_2}, as well as membranes \cite{Bucior_2012}, molecular motors \cite{Akimov_2012}, micelles \cite{Yan_2008}, and nanoparticles \cite{Patra_2014}.
The case of rigid water exemplifies this behavior and serves as an excellent test case for our proposal.
In fact, we were able to correct the discretization errors in numerically computed averages for time-step sizes up to $5~\text{fs}$, outperforming both the double-chain method by Kamberaj \textit{et al}. \cite{Kamberaj_2005} and a reimplementation of the single-chain method by Martyna \textit{et al}. \cite{Martyna_1996} for rigid bodies.

@@

In a previous paper \cite{Silveira_2017}, we took advantage of a particular factorization of the orientation matrix expressed in terms of quaternion components to
1) derive a new exact solution for torque-free rotations and use it as part of a symplectic integrator for rigid bodies and
2) reinterpret the approximate NO-SQUISH method \cite{Miller_2002} and unveil its equivalence to the earlier method by Dullweber and co-workers \cite{Dullweber_1997}.

@@

In these equations, $\vt r$ is the center-of-mass position of the body, $\vt p$ is its linear momentum, $\vt q$ is a unit quaternion that determines its orientation, and $\vt \pi$ is the quaternion-conjugate momentum.
Vectors $\vt F$ and $\vt \tau$ are, respectively, the resultant force and torque exerted on the body, both represented in the space-fixed frame of reference, and $\vt \omega = \frac{1}{2} {\mt I}^{-1} \tr{\mt B}(\vt q) {\vt \pi}$ is the three-dimensional angular velocity in the body-fixed frame.
Matrices $\mt M$ and $\mt I$ are diagonal ones.
The former contains the body mass in all three diagonal entries while the latter contains the three principal moments of inertia.
Finally, the matrix-valued functions $\mt B$ and $\vt C$ are defined as
\begin{equation*}
\label{eq:def_B_and_C}
\mt B(\vt q) = \begin{smallarray}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 & -q_4 &  q_3 \\
 q_4 &  q_1 & -q_2 \\
-q_3 &  q_2 &  q_1
\end{smallarray}
%
\; \text{and} \;
%
\mt C(\vt q) = \begin{smallarray}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 &  q_4 & -q_3 \\
-q_4 &  q_1 &  q_2 \\
 q_3 & -q_2 &  q_1
\end{smallarray}.
\end{equation*}

The factorization which played a central role in Ref.~\citenum{Silveira_2017} is ${\mt A}(\vt q) = \tr{\mt B}(\vt q) {\mt C}(\vt q)$, where ${\mt A}(\vt q)$ is the rotation matrix that connects the body-fixed and space-fixed frames. Eq.~\eqref{eq:ODE system for NVE} preserves the constraints $\tr{\vt q}{\vt q} = 1$ and $\tr{\vt q}{\vt \pi} = 0$, as well as the value of a Hamiltonian
\begin{equation}
\label{eq:Hamiltonian}
\Ham{} = K(\vt p, \vt q, \vt \pi) + U(\vt r,\vt q),
\end{equation}
where $K$ and $U$ are the kinetic and potential energies of the body, respectively.
While $U(\vt r, \vt q)$ is an arbitrary function, the kinetic energy is given by
\begin{equation}
\label{eq:kinetic energy}
K = \frac{1}{2} \tr{\vt p} {\mt M}^{-1} {\vt p} + \frac{1}{8} \tr{\vt \pi} {\mt B}(\vt q) {\mt I}^{-1} \tr{\mt B}(\vt q) {\vt \pi}.
\end{equation}

@@

A symplectic, time-reversible integrator can be devised by splitting the exponential operator according to the Trotter-Suzuki formula \cite{Trotter_1959, Suzuki_1976}
\begin{equation}
\label{eq:splitting NVE}
e^{\timestep \Liu{\tiny NVE}} = e^{\frac{\timestep}{2} \Liu{B}} e^{\timestep \Liu{A}} e^{\frac{\timestep}{2} \Liu{B}} + \mathcal{O}(\timestep^3),
\end{equation}
where $\Liu{A} + \Liu{B} = \Liu{\tiny NVE}$.
In the Verlet-type splitting of Ref.~\citenum{Silveira_2017}, the action of propagator $e^{\frac{\timestep}{2} \Liu{B}}$ is a kick in the linear and quaternion momenta given by ${\vt p} = {\vt p}^\ast + \frac{\timestep}{2} {\vt F}^\ast$ and ${\vt \pi} = {\vt \pi}^\ast + \timestep {\mt C}({\vt q^\ast}) {\vt \tau}^\ast$, respectively, where the asterisk denotes the state of the system immediately before the propagation occurs.
The action of $e^{\timestep \Liu{A}}$ is a simultaneous uniform translation (${\vt r} = {\vt r}^\ast + \timestep {\mt M}^{-1} {\vt p}^\ast$) and torque-free rotation along a full time step.
We have derived \cite{Silveira_2017} an exact solution for torque-free rotations in a form that facilitates computer implementation and allows, differently from other existing solutions \cite{Kosenko_1998, vanZon2007, Celledoni_2008}, a unified treatment of asymmetric, symmetric, and spherical tops.
This solution was used to evaluate the simpler NO-SQUISH method \cite{Miller_2002}.
Despite its approximate treatment of rotations, which are split into several revolutions around the principal axes \cite{Dullweber_1997}, that method proved to be very accurate in liquid-phase simulations \cite{Silveira_2017}.
This justifies its more widespread use.
However, having an unplit solution for rotations at hand is crucial for the feasibility of backward error analysis, as it might become clear in Sec.~\ref{sec:backward error analysis} and Appendix \ref{sec:rigid body shadow hamiltonian}.

@@

With $K$ defined as in Eq.~\eqref{eq:kinetic energy}, it turns out that $G_1 = 2K - L k_\text{\tiny B} T$.
If there are no external forces, the vector quantity $e^{\eta_1}\sum_{i=1}^N {\vt p}_i$ and the extended energy $H$ are conserved, while the system's center-of-mass position and the $\eta$ coordinates can be considered as a driven variables \cite{Tuckerman_2001}.
Moreover, $2N$ equations must be eliminated due to the constraints involving $\vt q$ and $\vt \pi$.
In the general case, by taking all these facts into account and applying the analysis of Tuckerman \textit{et al}. \cite{Tuckerman_2001}, we can deduce that the correct canonical distribution is attained if we make $L = 6N$.
In a particular situation in which we set $\sum_{i=1}^N {\vt p}_i = \vt 0$ at the onset of the simulation, we must make $L = 6N - 3$ \cite{Martyna_1994}.

Eq.~\eqref{eq:ODE system for NVT} defines an invariant measure in the extended phase-space \cite{Tuckerman_1999}, and it is possible to devise numerical solvers that preserve such measure \cite{Sergi_2001, Ezra_2004, Ezra_2006}.
We have done it by adapting the integrator introduced by Martyna \textit{et al}. \cite{Martyna_1996} in light of a criterion explained by Ezra \cite{Ezra_2006}.
It is worth noting, however, that such adaptation required us to alter the integration of the $\eta$ coordinates only.
Since these are driven variables, with no influence on the dynamics of the system, we can conclude that the practical importance of such alteration is small.

@@

The observation that $\refined{\Ham{}}$ still has a quadratic-form dependency on the momenta has an important consequence.
It suffices for the validity of Tolman's generalized equipartition theorem \cite{Tolman_1918, Uline_2008, Eastwood_2010} regarding a system with this Hamiltonian.
Moreover, as such dependency lies exclusively in $\refined K$, an outcome of this theorem for a system with $L$ degrees of freedom is that
\begin{equation}
\label{eq:equipartition}
L k_\text{\tiny B} T = \left\langle \tr{\vt p} \diff{\refined K}{\vt p} + \tr{\vt \pi} \diff{\refined K}{\vt \pi} \right\rangle_\timestep = 2\langle \refined{K} \rangle_\timestep,
\end{equation}
where subscript $\timestep$ points out the dependency of this average on the employed time step size.
The expression above extends the main result of Eastwood \textit{et al}. \cite{Eastwood_2010} to systems containing rigid bodies.
It provides a temperature estimator in substitution to the ordinary one, which is proportional to $\langle K \rangle$.
We emphasize that such temperature concerns the system that has actually been simulated, rather than the one initially specified.

@@

Bringing the attention back to the NVT case, our proposal for dealing with discretization effects goes as follows.
It consists in acknowledging that the middle propagator in Eq.~\eqref{eq:splitting NVT}, if split according to Eq.~\eqref{eq:splitting NVE}, will produce a phase-space move that is more closely consistent with the modified Hamiltonian equations than with the original ones.
Thus, we should adjust the Nos\'e-Hoover chain with the aim of obtaining a distribution proportional to $e^{-\beta \refined{\Ham{}}}$ rather than $e^{-\beta \Ham{}}$, where $\beta = \frac{1}{k_\text{\tiny B} T}$.
Fortunately, this entails solving Eq.~\eqref{eq:ODE system for NVT} almost exactly, by means of splitting, as described before.
The only required modification is that $\refined K$ should replace $K$ in Eq.~\eqref{eq:generalized force NVT}, which then becomes
\begin{equation}
\label{eq:thermostat_force}
G_1 = 2 \refined K - L k_\text{\tiny B} T.
\end{equation}

@@

\begin{figure}
\begin{tikzpicture}
\draw[->, dashed]   (0,0.0) -- (7,0.0) node[right]{${\vt r}(t)$};
\node(0) at (0,0)[shape=circle,draw,double,scale=0.5]        {};
\node(1) at (2.2,0)[fill=black,shape=circle,draw,scale=0.5]        {};
\node(2) at (4.4,0)[fill=black,shape=circle,draw,scale=0.5]        {};
\node(3) at (6.6,0)[shape=circle,draw,double,scale=0.5] {};
\node(4) at (3.4,0.8)[] {$e^{\timestep \Liu{\tiny NVE}}$};
\node at (0.0,-0.2)[below]{${\vt r}_{t-2\timestep}$};
\node at (2.2,-0.2)[below]{${\vt r}_{t-\timestep}$};
\node at (4.4,-0.2)[below]{${\vt r}_{t}$};
\node at (6.6,-0.2)[below]{${\vt r}_{t+\timestep}$};
\path[->,thick] (1) edge [bend left=45] (2);
\path[dashed,->,thick,postaction={decorate,decoration={raise=2.5ex,text along path,text align=center,text={|\small|virtual step}}}] (0) to [bend left=45] (1);
\path[dashed,->,thick] (1) edge [bend right=45] (0);
\path[dashed,->,thick,postaction={decorate,decoration={raise=2.5ex,text along path,text align=center,text={|\small|virtual step}}}] (2) to [bend left=45] (3);
\path[dashed,->,thick] (2) edge [bend left=45] (3);
\end{tikzpicture}
\caption{Schematic representation}
\label{fig:virtual steps}
\end{figure}

@@

\subsection{Refinement of Stochastic NVT Dynamics}
\label{sec:refined_bussi}

@@

In the previous section, we have seen that drifts in the extended energy of an NVT simulation can accumulate in the thermostat-related variables, thus preventing full-fledged instability.
This allows simulations to be executed with large time steps, which could in principle be used to save computer time.
However, as we will show in this section, discretization errors can become manifest much earlier than any risk of instability.
Fortunately, our proposed methodology is capable of mitigating the effect of these errors on ensemble averages computed for the simulated system.

In Fig.~\ref{fig:properties}, we present some thermodynamic properties computed for our liquid-water system using several NVT-MD methods with various time-step sizes.
All simulations were carried out for $T = 298.0~\text{K}$ and the analyzed quantities are the estimated temperature $T_\text{est} = \frac{2\langle K \rangle}{(6N-3)k_\mini{B}}$,
the mean potential energy per molecule $\langle U/N \rangle$,
the mean internal virial per molecule $\langle W/N \rangle$, and
the constant-volume specific heat $c_\mini{V} = \frac{\langle \Ham{}^2 \rangle - \langle \Ham{} \rangle^2}{N k_\mini{B} T^2}$.
The internal virial $W$ is computed as described in Ref.~\citenum{Silveira_2017}.
All symbols in Fig.~\ref{fig:properties} represent results obtained after $18~\text{ns}$ of production time, while the lines are just guides for the eyes.
The horizontal, dotted lines are meant to help visualizing the departure of the computed ensemble averages from their expected values.
They correspond to average results obtained with $\timestep = 1~\text{fs}$ by employing the three refined methods (details below).
Error bars represent 95\% confidence intervals obtained by using the Overlapping Batch Means estimator \cite{Meketon_1984, Flegal_2010} and standard uncertainty propagation formulas.

\begin{figure*}
	\includegraphics{Figures/thermodynamic_properties.eps}
	\caption{Effect of the time-step size on the thermodynamic properties and of 903 TIP3P\cite{Jorgensen_1983} water molecules in NVT MD simulations employing...}
	\label{fig:properties}
\end{figure*}

Fig.~\ref{fig:properties} contains results of five different NVT integrators.
Our measure-preserving extension to rigid bodies of the splitting scheme by Martyna \textit{et al}. \cite{Martyna_1996} is the basis of three of them.
Recall that this is a deterministic method in which a single Nos\'{e}-Hoover chain thermostat \cite{Martyna_1992} is attached to all degrees of freedom.
First, the method is applied in a conventional, refinement-free way, with the unsplit solution of Ref.~\citenum{Silveira_2017} being used for handling rotations.
Then, it is modified by introducing the refinement/reweighting procedure proposed in Sec.~\ref{sec:refined_method}.
In Fig.~\ref{fig:properties}, these two integrators are referred to as ``Martyna'' and ``Refined/NHC'', respectively.
Next, we replace the unsplit rotation solution by the simpler NO-SQUISH splitting of Miller \textit{et al}. \cite{Miller_2002}, denoting the resulting integrator by ``Refined/Miller'' in Fig.~\ref{fig:properties}.
We note that our refinement procedure is rigorously sustained (see Appendix \ref{sec:rigid body shadow hamiltonian}) for unsplit rotations only.
Nevertheless, it is known that NO-SQUISH rotations approach them almost indistinguishably in the case of water if small time steps are employed \cite{Silveira_2017}.
The fourth integrator we test is a refined version of the velocity-rescaling method of Bussi \textit{et al}. \cite{Bussi_2007}, which differs from the others for involving a single stochastic thermostat.
This is identified in Fig.~\ref{fig:properties} as ``Refined/Bussi''.
Finally, we also consider the double-thermostat integrator of Kamberaj \textit{et al}. \cite{Kamberaj_2005} for comparison purposes, despite the spurious energy flux between thermostats described in the preceding section.
This is done to show that the method, denoted by ``Kamberaj'' in Fig.~\ref{fig:properties}, also exhibits discretization-related issues not present in other integrators.

Before discussing the results shown in Fig.~\ref{fig:properties}, we stress that the thermodynamic properties are estimated in distinct ways for conventional and refined integrators.
For the conventional ones, Martyna and Kamberaj, simple averages are taken from the sampled configurations.
For the refined ones, reweighting is carried out via Eq.~\eqref{eq:reweighting}.

Fig.~\ref{fig:properties}(a)

>>>>

It seems reasonable to argue that any thermostat should regulate the shadow kinetic energy instead of the specified one.
However, the theory underlying the concept of shadow Hamiltonian is only applicable to symplectic integrators.
In this paper, we devise a novel procedure that partially overcomes this limitation.
It consists in using a splitting-based numerical integrator that entails a Hamiltonian component and applying backward error analysis only to this part of the integrator.
This allows us to obtain, with minimal computational overhead, a \textit{refined kinetic energy} and a \textit{refined Hamiltonian} as suitable estimators of their shadow counterparts.
By recasting the thermostat equations with this refined kinetic energy, we are able to predict its equipartition and to reproduce a canonical ensemble consistent with the refined Hamiltonian.
Finally, a reweighting procedure provides us with ensemble averages which are, instead, consistent with the specified one.
Although we will show the detailed derivation for the particular case of the Nos\'{e}-Hoover chain thermostat, it should be emphasized that our new approach is not limited to deterministic thermostats. 
It is straightforward to use it, for instance, in conjunction with global stochastic thermostats such as those of Bussi \textit{et al}. \cite{Bussi_2007} and Leimkuhler \textit{et al}. \cite{Leimkuhler_2009}.

Our new method is particularly appealing for rigid-body MD, primarily because its numerical stability extends up to time-step sizes large enough to yield pronounced discretization errors.
The case of rigid water exemplifies this behavior and serves as an excellent test case for our proposal.
In fact, we were able to correct the discretization errors in numerically computed averages for time-step sizes up to $5~\text{fs}$, outperforming both the double-chain method by Kamberaj \textit{et al}. \cite{Kamberaj_2005} and a reimplementation of the single-chain method by Martyna \textit{et al}. \cite{Martyna_1996} for rigid bodies.

It should be pointed out that rigid-body MD is relevant not only because some small molecules are usually modeled as rigid, such as the ubiquitous case of water \cite{Jorgensen_1983}, but also because it can provide a coarse-graining strategy for molecules that can be designed as collections of interconnected bodies, such as proteins and nanoparticles \cite{Miller_2002, Knorowski_2012, Patra_2013}.
In fact, the approach of Kamberaj \textit{et al}. \cite{Kamberaj_2005}, which is implemented in software packages such as HOOMD-\textit{blue}\cite{Anderson_2008} and LAMMPS\cite{Plimpton_1995}, has been applied to simulate small molecules \cite{Geiger_2013, Aimoli_2014, Aimoli_2014_2}, as well as membranes \cite{Bucior_2012}, molecular motors \cite{Akimov_2012}, micelles \cite{Yan_2008}, and nanoparticles \cite{Patra_2014}.

@@

In a previous paper \cite{Silveira_2017}, we took advantage of a particular factorization of the orientation matrix expressed in terms of quaternion components to derive a new exact solution for torque-free rotations and use it as part of a symplectic integrator for rigid bodies.

@@

In these equations, $\vt r$ is the center-of-mass position of the body, $\vt p$ is its linear momentum, $\vt q$ is a unit quaternion that determines its orientation, and $\vt \pi$ is the quaternion-conjugate momentum.
Vectors $\vt F$ and $\vt \tau$ are, respectively, the resultant force and torque exerted on the body, both represented in the space-fixed frame of reference, and $\vt \omega = \frac{1}{2} {\mt I}^{-1} \tr{\mt B}(\vt q) {\vt \pi}$ is the three-dimensional angular velocity in the body-fixed frame.
Matrices $\mt M$ and $\mt I$ are diagonal ones.
The former contains the body mass in all three diagonal entries while the latter contains the three principal moments of inertia.
Finally, the matrix-valued functions $\mt B$ and $\vt C$, given by
\begin{equation*}
\label{eq:def_B_and_C}
\mt B(\vt q) = \begin{smallarray}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 & -q_4 &  q_3 \\
 q_4 &  q_1 & -q_2 \\
-q_3 &  q_2 &  q_1
\end{smallarray}
%
\; \text{and} \;
%
\mt C(\vt q) = \begin{smallarray}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 &  q_4 & -q_3 \\
-q_4 &  q_1 &  q_2 \\
 q_3 & -q_2 &  q_1
\end{smallarray},
\end{equation*}
are related to the rotation matrix ${\mt A}(\vt q)$ through the factorization ${\mt A}(\vt q) = \tr{\mt B}(\vt q) {\mt C}(\vt q)$ that was mentioned at the beginning of this section. Eq.~\eqref{eq:ODE system for NVE} preserves the constraints $\tr{\vt q}{\vt q} = 1$ and $\tr{\vt q}{\vt \pi} = 0$, as well as the value of a Hamiltonian
\begin{equation}
\label{eq:Hamiltonian}
\Ham{} = K(\vt p, \vt q, \vt \pi) + U(\vt r,\vt q),
\end{equation}
where $K$ and $U$ are the kinetic and potential energies of the body, respectively.
While $U(\vt r, \vt q)$ is an arbitrary function, the kinetic energy is given by
\begin{equation}
\label{eq:kinetic energy}
K = \frac{1}{2} \tr{\vt p} {\mt M}^{-1} {\vt p} + \frac{1}{8} \tr{\vt \pi} {\mt B}(\vt q) {\mt I}^{-1} \tr{\mt B}(\vt q) {\vt \pi}.
\end{equation}

@@

A symplectic, time-reversible integrator can be devised by splitting the exponential operator according to the Trotter-Suzuki formula \cite{Trotter_1959, Suzuki_1976}
\begin{equation}
\label{eq:splitting NVE}
e^{\timestep \Liu{\tiny NVE}} = e^{\frac{\timestep}{2} \Liu{B}} e^{\timestep \Liu{A}} e^{\frac{\timestep}{2} \Liu{B}} + \mathcal{O}(\timestep^3),
\end{equation}
where $\Liu{A} + \Liu{B} = \Liu{\tiny NVE}$.
In the Verlet-type splitting of Ref.~\citenum{Silveira_2017}, the action of propagator $e^{\frac{\timestep}{2} \Liu{B}}$ is a kick in the linear and quaternion momenta given by ${\vt p} = {\vt p}^\ast + \frac{\timestep}{2} {\vt F}^\ast$ and ${\vt \pi} = {\vt \pi}^\ast + \timestep {\mt C}({\vt q^\ast}) {\vt \tau}^\ast$, respectively; where the asterisk denotes the state of the system immediately before the propagation occurs.
The action of $e^{\timestep \Liu{A}}$ is a simultaneous uniform translation (${\vt r} = {\vt r}^\ast + \timestep {\mt M}^{-1} {\vt p}^\ast$) and torque-free rotation along a full time step.
We have derived \cite{Silveira_2017} an exact solution for torque-free rotations in a form that facilitates computer implementation and allows, differently from other existing solutions \cite{Kosenko_1998, vanZon2007, Celledoni_2008}, a unified treatment of asymmetric, symmetric, and spherical tops.
This solution was used to evaluate the simpler NO-SQUISH method \cite{Miller_2002}.
Despite its approximate treatment of rotations, which are split into several revolutions around the principal axes \cite{Dullweber_1997}, that method proved to be very accurate in liquid-phase simulations \cite{Silveira_2017}.
This justifies its more widespread use.
However, having an unplit solution for rotations at hand is crucial for the feasibility of backward error analysis, as it might become clear in Sec.~\ref{sec:backward error analysis} and Appendix \ref{sec:rigid body shadow hamiltonian}.

@@

With $K$ defined as in Eq.~\eqref{eq:kinetic energy}, it turns out that $G_1 = 2K - L k_\text{\tiny B} T$.
If there are no external forces, the vector quantity $e^{\eta_1}\sum_{i=1}^N {\vt p}_i$ and the extended energy $H$ are conserved, while the system's center-of-mass position and the $\eta$ coordinates can be considered as driven variables \cite{Tuckerman_2001}.
Moreover, $2N$ equations must be eliminated due to the constraints involving $\vt q$ and $\vt \pi$.
In the general case, by taking all these facts into account and applying the analysis of Tuckerman \textit{et al}. \cite{Tuckerman_2001}, we can deduce that the correct canonical distribution is attained if we make $L = 6N$.
In a particular situation in which we set $\sum_{i=1}^N {\vt p}_i = \vt 0$ at the onset of the simulation, we must make $L = 6N - 3$ \cite{Martyna_1994}.

Eq.~\eqref{eq:ODE system for NVT} defines an invariant measure in the extended phase-space \cite{Tuckerman_1999}, and it is possible to devise numerical solvers that preserve such measure \cite{Sergi_2001, Ezra_2004, Ezra_2006}.
We have done it by adapting the integrator introduced by Martyna \textit{et al}. \cite{Martyna_1996} in light of a criterion explained by Ezra \cite{Ezra_2006}.
It is worth noting, however, that such adaptation required us to only alter the integration of the $\eta$ coordinates.
Since these are driven variables, with no influence on the dynamics of the system, we can conclude that the practical importance of such alteration is small.

@@

The observation that $\refined{\Ham{}}$ still has a quadratic-form dependency on the momenta has an important consequence: it ensures
the validity of Tolman's generalized equipartition theorem \cite{Tolman_1918, Uline_2008, Eastwood_2010} regarding a system with this Hamiltonian.
Moreover, as such dependency lies exclusively in $\refined K$, an outcome of this theorem for a system with $L$ degrees of freedom is that
\begin{equation}
\label{eq:equipartition}
L k_\text{\tiny B} T = \left\langle \tr{\vt p} \diff{\refined K}{\vt p} + \tr{\vt \pi} \diff{\refined K}{\vt \pi} \right\rangle_\timestep = 2\langle \refined{K} \rangle_\timestep,
\end{equation}
where subscript $\timestep$ points out the dependency of this average on the employed time step size.
The expression above extends the main result of Eastwood \textit{et al}. \cite{Eastwood_2010} to systems containing rigid bodies.
It provides a temperature estimator in substitution to the ordinary one, which is proportional to $\langle K \rangle$.
We emphasize that the temperature in Eq.~\ref{eq:equipartition} concerns the system that has actually been simulated, rather than the one initially specified.

@@

Bringing the attention back to the NVT case, our proposal for dealing with discretization effects consists in acknowledging that the middle propagator in Eq.~\eqref{eq:splitting NVT}, if split according to Eq.~\eqref{eq:splitting NVE}, will produce a phase-space move that is more closely consistent with the modified Hamiltonian equations than with the original ones.
Thus, we should adjust the Nos\'e-Hoover chain with the aim of obtaining a distribution proportional to $e^{-\beta \refined{\Ham{}}}$ rather than $e^{-\beta \Ham{}}$, where $\beta = \frac{1}{k_\text{\tiny B} T}$.
Fortunately, this entails solving Eq.~\eqref{eq:ODE system for NVT} almost exactly, by means of splitting, as described before.
The only required modification is that $\refined K$ should replace $K$ in Eq.~\eqref{eq:generalized force NVT}, which then becomes
\begin{equation}
\label{eq:thermostat_force}
G_1 = 2 \refined K - L k_\text{\tiny B} T.
\end{equation}

@@

\begin{figure}
\begin{tikzpicture}
\draw[->, dashed]   (0,0.0) -- (7,0.0) node[right]{${\vt r}(t)$};
\node(0) at (0,0)[shape=circle,draw,double,scale=0.5]        {};
\node(1) at (2.2,0)[fill=black,shape=circle,draw,scale=0.5]        {};
\node(2) at (4.4,0)[fill=black,shape=circle,draw,scale=0.5]        {};
\node(3) at (6.6,0)[shape=circle,draw,double,scale=0.5] {};
\node(4) at (3.4,0.8)[] {$e^{\timestep \Liu{\tiny NVE}}$};
\node at (0.0,-0.2)[below]{${\vt r}_{t-2\timestep}$};
\node at (2.2,-0.2)[below]{${\vt r}_{t-\timestep}$};
\node at (4.4,-0.2)[below]{${\vt r}_{t}$};
\node at (6.6,-0.2)[below]{${\vt r}_{t+\timestep}$};
\path[->,thick] (1) edge [bend left=45] (2);
\path[dashed,->,thick,postaction={decorate,decoration={raise=2.5ex,text along path,text align=center,text={|\small|virtual step}}}] (0) to [bend left=45] (1);
\path[dashed,->,thick] (1) edge [bend right=45] (0);
\path[dashed,->,thick,postaction={decorate,decoration={raise=2.5ex,text along path,text align=center,text={|\small|virtual step}}}] (2) to [bend left=45] (3);
\path[dashed,->,thick] (2) edge [bend left=45] (3);
\end{tikzpicture}
\caption{Schematic representation of the executed steps in a single time-step of the refined NVT methods.
The solid arrow represents the NVE part of the integration, while the dotted arrows emphasize that forward and backward steps are carried out for positions and orientations with the only purpose of approximating their time-derivatives at point $t$.}
\label{fig:virtual steps}
\end{figure}

@@

@@

In the previous section, we have seen that drifts in the extended energy of an NVT simulation can accumulate in the thermostat-related variables, thus preventing numerical instability.
This allows simulations to be executed with large time steps, which could in principle be used to save computer time.
However, as we will show in this section, discretization errors can become manifest much earlier than any risk of instability.
Fortunately, our proposed methodology is capable of mitigating the effect of these errors on ensemble averages computed for the simulated system.

In Fig.~\ref{fig:properties}, we present some thermodynamic properties computed for our liquid-water system using several NVT-MD methods with various time-step sizes.
All simulations were carried out for $T = 298.0~\text{K}$ and the analyzed quantities are the estimated temperature $T_\text{est} = \frac{2\langle K \rangle}{(6N-3)k_\mini{B}}$,
the mean potential energy per molecule $\langle U/N \rangle$,
the mean internal virial per molecule $\langle W/N \rangle$, and
the constant-volume specific heat $c_\mini{V} = \frac{\langle \Ham{}^2 \rangle - \langle \Ham{} \rangle^2}{N k_\mini{B} T^2}$.
The internal virial $W$ is computed as described in Ref.~\citenum{Silveira_2017}.
All symbols in Fig.~\ref{fig:properties} represent results obtained after $18~\text{ns}$ of production time, while the lines are just guides for the eyes.
The horizontal dotted lines are meant to help visualize the departure of the computed ensemble averages from their expected values, which correspond to average results obtained with $\timestep = 1~\text{fs}$ by employing the three refined methods (details below).
Error bars represent 95\% confidence intervals obtained by using the Overlapping Batch Means estimator \cite{Meketon_1984, Flegal_2010} and standard uncertainty propagation formulas.

\begin{figure*}
	\includegraphics{Figures/thermodynamic_properties.eps}
	\caption{Effect of the time-step size on (a) the estimated temperature, (b) the mean potential energy per molecule, (c) the mean internal virial per molecule, and (d)
the constant-volume specific heat, as obtained for a system of 903 TIP3P\cite{Jorgensen_1983} water molecules in NVT MD simulations employing different numerical schemes: ``Martyna'' corresponds to our implementation for rigid bodies of the Nos\'{e}-Hoover chain thermostat introduced in Ref.~\citenum{Martyna_1996} (red circles), the ``Refined/NHC'' scheme (black squares) is obtained by refining ``Martyna'', as described in Sec.~\ref{sec:refined_method}, ``Refined/NHC/Miller'' (pink up triangles) differs from this last one in that it employs the approximate solution of free rotations\cite{Miller_2002}, ``Refined/Bussi'' (green x) stands for a refined version of the stochastic method of Bussi \textit{et al}. \cite{Bussi_2007}  and, finally, ``Kamberaj'' (blue down triangles) refers to the double-thermostat introduced by Kamberaj \textit{et al.} \cite{Kamberaj_2005}  .}
	\label{fig:properties}
\end{figure*}

Fig.~\ref{fig:properties} contains results of five different NVT integrators.
Our measure-preserving extension to rigid bodies of the splitting scheme by Martyna \textit{et al}. \cite{Martyna_1996} is the basis of three of them.
Recall that this is a deterministic method in which a single Nos\'{e}-Hoover chain thermostat \cite{Martyna_1992} is attached to all degrees of freedom.
First, the method is applied in a conventional, refinement-free way, with the unsplit solution of Ref.~\citenum{Silveira_2017} being used for handling rotations.
Then, it is modified by introducing the refinement/reweighting procedure proposed in Sec.~\ref{sec:refined_method}.
In Fig.~\ref{fig:properties}, these two integrators are referred to as ``Martyna'' and ``Refined/NHC'', respectively.
Next, we replace the unsplit rotation solution by the simpler NO-SQUISH splitting of Miller \textit{et al}. \cite{Miller_2002}, denoting the resulting integrator by ``Refined/Miller'' in Fig.~\ref{fig:properties}.
We note that our refinement procedure is rigorously valid (see Appendix \ref{sec:rigid body shadow hamiltonian}) for unsplit rotations only.
Nevertheless, it is known that NO-SQUISH rotations approach them almost indistinguishably in the case of water if small time steps are employed \cite{Silveira_2017}.
The fourth integrator we test is a refined version of the velocity-rescaling method of Bussi \textit{et al}. \cite{Bussi_2007}, which differs from the others for involving a single stochastic thermostat.
This is identified in Fig.~\ref{fig:properties} as ``Refined/Bussi''.
Finally, we also consider the double-thermostat integrator of Kamberaj \textit{et al}. \cite{Kamberaj_2005} for comparison purposes, despite the spurious energy flux between thermostats described in the preceding section.
This is done to show that the method, denoted by ``Kamberaj'' in Fig.~\ref{fig:properties}, also exhibits discretization-related issues not present in other integrators.

Before discussing the results shown in Fig.~\ref{fig:properties}, we stress that the thermodynamic properties are estimated in distinct ways for conventional and refined integrators.
For the conventional ones, ``Martyna'' and ``Kamberaj'', simple averages are taken from the sampled configurations.
For the refined ones, reweighting is carried out via Eq.~\eqref{eq:reweighting}.

In Fig.~\ref{fig:properties}(a), we see that the estimated temperature for ``Kamberaj'' decreases appreciably with increasing $\timestep$.
This behavior is typical of Verlet-like NVE integrators\cite{Davidchack_2010, Silveira_2017}, reflecting that particle momenta is weakly modified by the NHC factorization scheme of Kamberaj \textit{et al}. \cite{Kamberaj_2005}.
On the other hand, the stringent temperature control achieved in ``Martyna'' leads to a severely degraded accuracy in both $\langle U \rangle$ and $\langle W/N \rangle$, as noted in Figs.~\ref{fig:properties}(b)-(c).
According to the results of Fig.~\ref{fig:nve}, discretization errors in the NVE integration become manifest at $2~\text{fs}$.
Thus, the apparent success of the scheme by Martyna \textit{et al.} \cite{Martyna_1996}, in terms of the kinetic energy control for the entire range of time-step sizes considered, is wiped out by those discretization errors which, if not properly handled, yield considerably degraded configurational properties.
